\huge \textbf{Parameter Forecasts From CMB Lensing and Galaxy Lensing Power- and Bispectra}

### 2025-05-15 paper/paper_draft.tex
paper changes
misc

\author[1]{\textbf{Jonas Frugte}}
\author[2]{\textbf{P. Daniel Meerburg}}

\affil[1]{Independent Researcher, The Netherlands}
\affil[2]{Van Swinderen Institute for Particle Physics and Gravity, University of Groningen, Nijenborgh 4, 9747 AG Groningen, The Netherlands}
\affil[ ]{}
\affil[ ]{\textit{E-mail:} \texttt{jonasfrugte@gmail.com}}
 % Removes the default date

@@

Weak lensing has been used as a probe to constrain cosmological parameters since the 2000's \cite{kilbinger2015cosmology,Bacon2000,Kaiser2000,Waerbeke2000,Bartelmann2001}. A range of upcoming surveys, such as the survey by the Simons Observatory (SO) \cite{Ade2019}, the Legacy Survey of Space and Time (LSST) \cite{Ivezic2019}, and Euclid \cite{Laureijs2011} aim to measure weak lensing of the CMB and of galaxies. They will do so much more accurately than previous surveys, enabling tighter constraints on parameters and progressing our understanding of physics as a whole. Typically, the main value of interest is the power spectrum (or equivalently, two point function) of the lensing potential. This can be measured in terms of the lensing convergence $\kappa$, lensing shear $\gamma$, or lensing potential $\psi$. In the weak lensing regime they all contain the same amount of information and can be easily converted into one another, see section \ref{sec:weaklensstats}. For the CMB the lensing potential is generally measured directly \cite{Planck2018Lensing}, while with galaxies one instead measures the shear, as this can be deduced from the ellipiticity distribution of the observed galaxies \cite{HoekstraJain2008}.

Even with Gaussian initial conditions, there are significant amounts of non-Gaussianity in the weak lensing signal, especially in galaxy lensing due to it being a probe of the more recent distribution of matter \cite{Bernardeau1997,Takada2003}. In addition to looking at the power spectrum, a natural next step is thus to measure the lensing bispectrum. This has already been done for signals from galaxies \cite{vanWaerbeke2002}, but no proper detection has been done of the CMB lensing bispectrum. Upcoming surveys such as the ones mentioned earlier may detect this signal. Therefore it is relevant to know the detectability of, and parameter constraints possible from, all 4 signals: CMB lensing power- and bispectra and galaxy lensing power- and bispectra. Notably, this includes looking at cross correlation as well.

In this article we aim to answer this question in the context of experimental parameters similar to those of next generation surveys such as mentioned earlier. We are especially interested in seeing if approximate parameter degeneracies can be broken by combining all data. Calculating the covariance matrix of unbiased optimal estimators of the cosmological parameters can be done by combining cosmological simulations (in our case we use the CAMB package \cite{Lewis2000}) with a Fisher matrix analysis. See figure \ref{fig:flowchart} for an overview of the calculation done in this paper. The parameters looked at are the Hubble constant, $H_0$, physical baryon density, $\Omega_b h^2$, physical cold dark matter density, $\Omega_c h^2$, scalar spectral index, $n_s$, amplitude of primordial scalar fluctuations, $A_s$, sum of neutrino masses $\sum m_\nu$, and dark energy equation of state parameter, $w_0$. Though, we will find that some of the parameters are poorly constrained even when combining all 4 signals and thus omit them from the analysis in the results. As a secondary purpose, this paper also includes a derivation for the Fisher matrix formalism in the non-trivial context of bispectra with multiple tracers. The final formula can already be found in the literature (see e.g.\ the appendix of \cite{Kalaja_2021}), but the authors of this paper were not able to find a derivation and hence included it.

The structure of this paper is as follows: we first review the theory of weak lensing and weak lensing statistics in sections \ref{sec:weaklensing} and \ref{sec:weaklensstats}. We then provide technical details of the calculation in section \ref{sec:calcdetails}, such as choices for noise and numerical approximations made in the code. In section \ref{sec:results} we present the results, consisting of signal to noise ratios and parameter constraints for different combinations of the CMB and galaxy lensing power- and bispectra. We conclude in section \ref{sec:discussion}. Appendix \ref{sec:fisher} contains a short review of the Fisher matrix formalism and gives a technical derivation of the formula used to calculate each Fisher matrix element in the case of bispectra measurements with multiple tracers. Finally, appendix \ref{sec:shear} explains how to relate the shear to the lensing potential in spherical harmonic space.

@@

\section{Theory}
\subsection{Weak lensing spectra}
Radiation from cosmological objects is distorted due to gravitational lensing. In the vast majority of cases this lensing is a relatively weak effect and is thus referred to as weak lensing. Weak lensing is quantified through the deflection field $\mathbf d(\hat {\mathbf n})$ which equals the difference between the observed angle of a point on the sky and the true (unlensed) angle. This field is the gradient of the lensing potential $\psi(\hat {\mathbf n})$ given as a weighted integral over the mass distribution along the line of sight between earth and the observed point.

@@

To constrain cosmological parameters we can then look at the lensing potential of the CMB, $\psi_\CMB$, and of galaxy surveys, $\psi_\gal$. These are directly related to the matter power and bispectra as
\begin{empheq}[box=\fbox]{align*}
    C^{\psi_X\psi_Y}_l
    & = \frac{9}{l^4}\Omega_m^2H_0^4\int_0^{\chi_*} \chi^2 \d\chi a(\eta_0-\chi)^{-2}W_X(\chi)W_Y(\chi)C^\delta(l/\chi,\eta_0-\chi),\\
    B^{\psi_X\psi_Y\psi_Z}_{l_1l_2l_3} &= \sqrt{\frac{(2l_1 + 1)(2l_2 + 1)(2l_3 + 1)}{4\pi}} \begin{pmatrix} l_1 & l_2 & l_3 \\ 0 & 0 & 0 \end{pmatrix} \frac{27}{l_1^2l_2^2l_3^2}\Omega_m^3H_0^6\\
    & \quad \times \int \chi^2\d \chi a(\eta_0-\chi)^{-3}W_X(\chi)W_Y(\chi)W_Z(\chi)  B^\delta(\{l_i/\chi\}, \eta_0-\chi),
\end{empheq}
with $X, Y, Z \in \{\CMB, \gal\}$. $\Omega_m$ is the present day matter density parameter, $H_0$ is the present day Hubble constant, $a(\eta)$ is the scale factor, $\eta_0$ is the conformal time today, $\chi$ is the comoving radial distance, $\chi_*$ the distance to surface of last scattering, $P^\delta(k, \eta)$ the matter power spectrum, $B^\delta(k_1, k_2, k_3, \eta)$ the matter bispectrum, $W_X(\chi)$ the window function, and
$$
\begin{pmatrix}
    l_1&l_2&l_3 \\ m_1 & m_2 & m_3
\end{pmatrix}
$$ 
the Wigner 3-j symbol. The only factor that differs between CMB and galaxy lensing is the window function, defined as
\begin{equation*}
    W_X(\chi) = \int_\chi^{\infty}\d\chi' p_X(\chi')\frac{\chi'-\chi}{\chi'\chi},
\end{equation*}
with $p_X$ the radius distribution of the radiation source. For the CMB we take $p(\chi) = \delta(\chi - \chi_*)$. Notice that we can also look at cross correlation between CMB and galaxy lensing.

For a derivation of the lensing potential and its power- and bispectra, see appendices \ref{sec:weaklensing} and \ref{sec:weaklensstats}.

\subsection{Nonlinear matter bispectrum}
Determining the nonlinear matter powerspectrum can be done using numerical codes such as CAMB. The nonlinear matter bispectrum was calculated from the powerspectrum using a fitting formula based on perturbation theory in \cite{bispfit}. It is given by
\begin{equation}
    B^\delta(k_1, k_2, k_3, \chi) = 2 F_2(k_1, k_2, z) C^\delta(k_1, z) C^\delta(k_2, z) + \text{2 perm.}
\end{equation}
where \( C^\delta \) is the nonlinear matter power spectrum\footnote{Compare to the tree level bispectrum where we instead use the linear powerspectrum.}, and the kernel \( F_2 \) is modified from the tree level result with factors \( a(k, z) \), \( b(k, z) \), and \( c(k, z) \):
\begin{equation}
    F_2(k_1, k_2, z) = \frac{5}{7} a(k_1, z) a(k_2, z) + \frac{k_1^2 + k_2^2}{2 k_1 k_2} b(k_1, z) b(k_2, z) \cos \theta + \frac{2}{7} c(k_1, z) c(k_2, z) \cos^2 \theta.
\end{equation}
They are defined as:
\begin{equation}
    a(k, z) = \frac{1 + \sigma_8^{a_6}(z) \sqrt{0.7} Q(n_{\text{eff}}) (q^{a_1})^{n_{\text{eff}} + a_2}}{1 + (q^{a_1})^{n_{\text{eff}} + a_2}},
\end{equation}
\begin{equation}
    b(k, z) = \frac{1 + 0.2 a_3 (n_{\text{eff}} + 3) (q^{a_7})^{n_{\text{eff}} + 3 + a_8}}{1 + (q^{a_5})^{n_{\text{eff}} + 3.5 + a_8}},
\end{equation}
\begin{equation}
    c(k, z) = \frac{1 + \left[ \frac{4.5 a_4}{1.5 + (n_{\text{eff}} + 3)^4} \right] (q^{a_5})^{n_{\text{eff}} + 3 + a_9}}{1 + (q^{a_5})^{n_{\text{eff}} + 3.5 + a_9}}.
\end{equation}
Here, \( Q(n_{\text{eff}}) \) is given by:
\begin{equation}
    Q(x) = \frac{4 - 2x}{1 + 2^{x+1}}.
\end{equation}
The effective spectral index of the linear power spectrum is defined as:
\begin{equation}
    n_{\text{eff}} \equiv \frac{d \ln C^\delta_{\text{lin}}(k)}{d \ln k}.
\end{equation}
Additionally, \( q \) is given by:
\begin{equation}
    q = \frac{k}{k_{\text{NL}}},
\end{equation}
where \( k_{\text{NL}} \) is the scale at which nonlinearities become significant, satisfying:
\begin{equation}
    4 \pi k_{\text{NL}}^3 C^\delta_{\text{lin}}(k_{\text{NL}}, 0) = 1.
\end{equation}
The coefficients \( a_i \) are:
\begin{align*}
    a_1 &= 0.484, \quad a_2 = 3.740, \quad a_3 = -0.849, \quad a_4 = 0.392, \\
    a_5 &= 1.013, \quad a_6 = -0.575, \quad a_7 = 0.128, \quad a_8 = -0.722, \quad a_9 = -0.926.
\end{align*}

\subsection{Fisher matrix analysis}
SNR and parameter constraints are obtained through a Fisher matrix analysis as is standard. A full derivation of below statements can be found in appendix \ref{sec:fisher}. The auto and cross power spectrum Fisher matrix is given as
\begin{equation*}
    F_{\alpha\beta} = \sum_{l} \sum_{XY}\sum_{X'Y'}(2l+1)\partial_\alpha C_l^{XY} (C^{-1})_l^{XX'}(C^{-1})_l^{YY'} \partial_\beta C^{X'Y'}_l
\end{equation*}
where
\begin{equation*}
    C_{l} := \begin{pmatrix}
        C^{\psi_\kappa\psi_\kappa}_l & C^{\psi_\kappa\psi_\gamma}_l \\
        C^{\psi_\kappa\psi_\gamma}_l & C^{\psi_\gamma\psi_\gamma}_l
    \end{pmatrix}.
\end{equation*}
For the bispectra we instead get 
\begin{equation*}
    F_{\alpha\beta} = \sum_{l_1 \leq l_2 \leq l_3} \frac{\mathcal P _{l_1l_2l_3}}{6}
    \sum_{XYZ}\sum_{X'Y'Z'} 
    \partial_\alpha B^{X Y Z}_{l_1 l_2 l_3} 
    (C^{-1})^{X X'}_{l_1}
    (C^{-1})^{Y Y'}_{l_2}
    (C^{-1})^{Z Z'}_{l_3}
    \partial_\beta B^{X' Y' Z'}_{l_1 l_2 l_3},
\end{equation*}
with $\mathcal P_{l_1l_2l_3}$ defined as the number of distinct permutations that can be made with $l_1l_2l_3$. When only considering autospectra the formulas simplify to
\begin{equation*}
    F_{\alpha\beta} = \sum_{l} \sum_{XY}\sum_{X'Y'}\frac{2l+1}{2}\frac{\partial_{\alpha}C_l^{XX} \partial_\beta C^{XX}_l}{(C_l^{XX})^2},
\end{equation*}
and (for bispectra)
\begin{equation*}
    F_{\alpha\beta} = \sum_{l_1 \leq l_2 \leq l_3} \frac{\mathcal P _{l_1l_2l_3}}{6}
    \frac{\partial_\alpha B^{X X X}_{l_1 l_2 l_3} 
    \partial_\beta B^{XXX}_{l_1 l_2 l_3}}{C^{XX}_{l_1}C^{XX}_{l_2}C^{XX}_{l_3}}.
\end{equation*}

@@

% lmin, lmax, galaxy redshift distribution, fiducial values (!), noise levels, 
% for SO, CMB - S4, Euclid, LSST
We chose our fiducial cosmology in line with the results from the Planck mission \cite{planckresults}.
\begin{align*}
    H_0 &= 67.4 \, \text{km/s/Mpc}, && \text{(Hubble constant)} \\
    \Omega_b h^2 &= 0.0223, && \text{(Physical baryon density parameter)} \\
    \Omega_c h^2 &= 0.119, && \text{(Physical cold dark matter density parameter)} \\
    n_s &= 0.965, && \text{(Scalar spectral index)} \\
    A_s &= 2.13 \times 10^{-9}, && \text{(Amplitude of primordial scalar fluctuations)} \\
    \sum m_\nu &= 0.06 \, \text{eV}, && \text{(Sum of neutrino masses)} \\
    w_0 &= -1. && \text{(Dark energy equation of state parameter)}
\end{align*}

@@

This paper considers noise levels for stage 3 and stage 4 weak lensing surveys. Values can be found in table \ref{tab:noiselevels}. A comparison of noise power spectrum to lensing power spectrum can be found in figure \ref{fig:lpsplusnoise}.

@@

CMB noise is estimated through the quadratic estimator developed in \cite{cmblensingestimator}. We do not take temperature measurements into account, as these are subleading for stage 4 experiments (REF). This is similarly done in \cite{Namikawa_2016}. The parameters characterizing the noise levels are thus beam width, $\sigma$ and polarization white noise, $\Delta_P$. Galaxy lensing is determined by measuring lensing shear (see appendices \ref{sec:weaklensing} and \ref{sec:shear}). The noise in this measurement is dominated by scale independent shot noise and has associated noise power spectrum $N_l^{\text{shear}} = \sigma_{\text{rms}}^2 / n_g$. $n_g$ is the amount of galaxies observed per unit angle. The noise power spectrum for the lensing potential then equals
\begin{equation*}
    N_l^{\text{lens. potential}} = \frac{(l-1)l(l+1)(l+2)}{4}N_l^{\text{shear}}.
\end{equation*}
For parameter constraints we use stage 4 noise levels.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{/Users/jonasfrugte/Desktop/Research_Project/fisher_calc_weak_lensing/code/plots/spectraplusnoise.pdf}
    \caption{CMB (right) and galaxy (left) lensing potential power spectra compared to associated experimental noise. Current (stage 3) noise values are displayed as well as near future (stage 4) noi se values. The CMB lensing experiment does not measure temperature anisotropies, only polarization. CMB noise values are chosen in accordance with \cite{Namikawa_2016}. Shear noise values are chosen to be similar to e.g. Euclid measurements for stage 4 and e.g. KiDS for stage 3.}
    \label{fig:lpsplusnoise}
\end{figure}

@@

\section{Results}\label{sec:results}
\subsection{SNR}
Lensing power spectra are well detectable in nearly all cases. 
The signal to noise ratios of the CMB and galaxy lensing bispectra v.s. the maximum multipole moment measured can be found in figure \ref{fig:snrplots}. Shear bispectra can be measured well by both stage 3 and stage 4 experiments, while CMB lensing is only semi detectable by a stage 3 experiment but well detectable by stage 4 experiments.
%It can be seen that the CMB bispectrum signal to noise never exceeds 2 even for $l_{\max} = 2000$, meaning that this signal is not likely to be able to contribute to parameter constraints in future surveys. All other quantities should be well detectable for $l_{\max}\geq 500$.
\begin{figure}
    \includegraphics[width=\textwidth]{figures/snrplots.pdf}
    \caption{Signal to noise ratios for galaxy lensing (dashed) and CMB lensing (non-dashes) power- (in red) and bispectra (in blue). The $x$-axis labels the maximum multipole moment measured up to, starting from $l_{\min}=2$.}
    \label{fig:snrplots}
\end{figure}

\subsection{Parameter Constraints}
Parameter constraints for the full set of $\Lambda$ CDM parameters are shown in figure \ref{fig:paramconstraintsall}. The combination $H_0$, $\sigma_8$, and $\Omega_m$ is particularly well constrained, constraints for these parameters alone (i.e. marginalized over all other parameters) are found in figure \ref{fig:paramconstraintstight}.

>>>>

\author[ ]{\textbf{Jonas Frugte}}
\author[1]{\textbf{P. Daniel Meerburg}}

% \affil[1]{Independent Researcher, The Netherlands}
\affil[1]{Van Swinderen Institute for Particle Physics and Gravity, University of Groningen, Nijenborgh 4, 9747 AG Groningen, The Netherlands}
\affil[ ]{}
\affil[ ]{\textit{E-mail:} \texttt{jonasfrugte@gmail.com}}
 % Removes the default date

@@

Weak lensing has been used as a probe to constrain cosmological parameters since the 2000's \cite{kilbinger2015cosmology,Bacon2000,Kaiser2000,Waerbeke2000,Bartelmann2001}. A range of upcoming surveys, such as the survey by the Simons Observatory (SO) \cite{Ade2019}, the Legacy Survey of Space and Time (LSST) \cite{Ivezic2019}, and Euclid \cite{Laureijs2011} aim to measure weak lensing of the CMB and of galaxies. They will do so much more accurately than previous surveys, enabling tighter constraints on parameters and progressing our understanding of physics as a whole. Typically, the statistic of interest is the power spectrum (or equivalently, two point function) of the lensing potential. This can be measured in terms of the lensing convergence $\kappa$, lensing shear $\gamma$, or lensing potential $\psi$. In the weak lensing regime they all contain the same amount of information and can be trivially converted into one another, see appendix \ref{sec:weaklensstats}. In CMB surveys the temperature and polarization anisotropies are used to estimate the lensing potential \cite{Planck2018Lensing}, while with galaxies one instead calculates lensing effects by measuring the lensing shear, which can be deduced from the ellipiticity distribution of the observed galaxies \cite{HoekstraJain2008}.

Even with Gaussian initial conditions, there are significant amounts of non-Gaussianity in the weak lensing signal, especially in galaxy lensing due to it being a probe of the more recent distribution of matter \cite{Bernardeau1997,Takada2003}. In addition to looking at the power spectrum, a natural next step is thus to measure the lensing bispectrum. This has already been done for signals from galaxies \cite{vanWaerbeke2002}, but we do not yet have any proper detection of the CMB lensing bispectrum. Upcoming surveys such as the ones mentioned earlier may detect this signal. Therefore it is timely to consider the detectability, as well as ability to constrain cosmological parameter of CMB lensing power- and bispectra, galaxy lensing power- and bispectra, and cross correlations between the two weak lensing observables.

In this paper we aim to answer this question in the context of experimental parameters similar to those of next generation surveys. We are especially interested in seeing if approximate parameter degeneracies can be broken by combining all data. Calculating the covariance matrix of unbiased optimal estimators of the cosmological parameters can be done by combining a Boltzmann code for cosmological perturbations with a Fisher matrix analysis. See figure \ref{fig:flowchart} for an overview of the calculation done in this paper. We look at $\lambda$CDM parameters as well as the sum of neutrino masses $\sum m_\nu$ and dark energy equation of state parameter, $w_0$.

Usefull derivations relevant to this paper can be found in ose, this paper also includes a derivation for the Fisher matrix formalism in the non-trivial context of bispectra with multiple tracers. The final formula can already be found in the literature (see e.g.\ the appendix of \cite{Kalaja_2021}), but the authors of this paper were not able to find a derivation and hence included it.

The structure of this paper is as follows: we introduce relevant formulas for weak lensing statistics, nonlinear matter bispectrum, and fisher matrix formalism in section \ref{sec:theory}. We specify details such as the fiducial values of the parameters and noise models used in section \ref{sec:calcdetails}. Parameter constraints and signal to noise ratios are presented in section \ref{sec:results} and discussed in section \ref{sec:discussion}. The appendices review the derivations of the formulas for weak lensing power- and bispectra. We also derive the Fisher matrix formalism, in particular for the more involved case of measuring bispectra with multiple tracers.

@@

\section{Theory}
\label{sec:theory}
\subsection{Weak lensing spectra}
Radiation from cosmological objects is distorted due to gravitational lensing. In the vast majority of cases this lensing is a relatively weak effect and is thus referred to as weak lensing. Weak lensing is quantified through the deflection field $\mathbf d(\hat {\mathbf n})$ which equals the difference between the observed angle of a point on the sky and the true (unlensed) angle. This field is the gradient of the lensing potential $\psi(\hat {\mathbf n})$ given as a weighted integral over the mass distribution along the line of sight between earth and the observed point.

@@

To constrain cosmological parameters we can then look at the lensing potential of the CMB, $\psi_\CMB$, and of galaxy surveys, $\psi_\gal$. These are directly related to the matter power and bispectra as
\begin{empheq}[box=\fbox]{align*}
    C^{\psi_X\psi_Y}_l
    & = \frac{9}{\ell^4}\Omega_m^2H_0^4\int_0^{\chi_*} \chi^2 \d\chi a(\eta_0-\chi)^{-2}W_X(\chi)W_Y(\chi)C^\delta(\ell/\chi,\eta_0-\chi),\\
    B^{\psi_X\psi_Y\psi_Z}_{\ell_1\ell_2\ell_3} &= \sqrt{\frac{(2l_1 + 1)(2l_2 + 1)(2l_3 + 1)}{4\pi}} \begin{pmatrix} l_1 & l_2 & l_3 \\ 0 & 0 & 0 \end{pmatrix} \frac{27}{l_1^2l_2^2l_3^2}\Omega_m^3H_0^6\\
    & \quad \times \int \chi^2\d \chi a(\eta_0-\chi)^{-3}W_X(\chi)W_Y(\chi)W_Z(\chi)  B^\delta(\{l_i/\chi\}, \eta_0-\chi),
\end{empheq}
with $X, Y, Z \in \{\CMB, \gal\}$. A derivation can be found in appendices \ref{sec:weaklensing} and \ref{sec:weaklensstats}. $\Omega_m$ is the present day matter density parameter, $H_0$ is the present day Hubble constant, $a(\eta)$ is the scale factor, $\eta_0$ is the conformal time today, $\chi$ is the comoving radial distance, $\chi_*$ the distance to surface of last scattering, $C^\delta(k, \eta)$ the matter power spectrum, $B^\delta(k_1, k_2, k_3, \eta)$ the matter bispectrum, $W_X(\chi)$ the window function, and
$$
\begin{pmatrix}
    l_1&l_2&l_3 \\ m_1 & m_2 & m_3
\end{pmatrix},
$$ 
the Wigner 3-j symbol. The only factor that differs between CMB and galaxy lensing is the window function, defined as
\begin{equation*}
    W_X(\chi) = \int_\chi^{\infty}\d\chi' p_X(\chi')\frac{\chi'-\chi}{\chi'\chi},
\end{equation*}
with $p_X$ the radius distribution of the radiation source. For the CMB we take $p(\chi) = \delta(\chi - \chi_*)$.

\subsection{Nonlinear matter bispectrum}
Determining the nonlinear matter powerspectrum can be done using numerical codes such as CAMB. The nonlinear matter bispectrum was calculated from the powerspectrum using a fitting formula based on perturbation theory in \cite{bispfit}. It is given by
\begin{equation}
    B^\delta(k_1, k_2, k_3, \chi) = 2 F_2(k_1, k_2, z) C^\delta(k_1, z) C^\delta(k_2, z) + \text{2 perm},
\end{equation}
where \( C^\delta \) is the nonlinear matter power spectrum\footnote{Compare to the tree level bispectrum where we instead use the linear powerspectrum.}, and the kernel \( F_2 \) is modified from the tree level result with factors \( a(k, z) \), \( b(k, z) \), and \( c(k, z) \):
\begin{equation}
    F_2(k_1, k_2, z) = \frac{5}{7} a(k_1, z) a(k_2, z) + \frac{k_1^2 + k_2^2}{2 k_1 k_2} b(k_1, z) b(k_2, z) \cos \theta + \frac{2}{7} c(k_1, z) c(k_2, z) \cos^2 \theta.
\end{equation}
They are defined as:
\begin{equation}
    a(k, z) = \frac{1 + \sigma_8^{a_6}(z) \sqrt{0.7} Q(n_{\text{eff}}) (q^{a_1})^{n_{\text{eff}} + a_2}}{1 + (q^{a_1})^{n_{\text{eff}} + a_2}},
\end{equation}
\begin{equation}
    b(k, z) = \frac{1 + 0.2 a_3 (n_{\text{eff}} + 3) (q^{a_7})^{n_{\text{eff}} + 3 + a_8}}{1 + (q^{a_5})^{n_{\text{eff}} + 3.5 + a_8}},
\end{equation}
\begin{equation}
    c(k, z) = \frac{1 + \left[ \frac{4.5 a_4}{1.5 + (n_{\text{eff}} + 3)^4} \right] (q^{a_5})^{n_{\text{eff}} + 3 + a_9}}{1 + (q^{a_5})^{n_{\text{eff}} + 3.5 + a_9}}.
\end{equation}
Here, \( Q(n_{\text{eff}}) \) is given by:
\begin{equation}
    Q(x) = \frac{4 - 2x}{1 + 2^{x+1}}.
\end{equation}
The effective spectral index of the linear power spectrum is defined as:
\begin{equation}
    n_{\text{eff}} \equiv \frac{d \ln C^\delta_{\text{lin}}(k)}{d \ln k}.
\end{equation}
Additionally, \( q \) is given by:
\begin{equation}
    q = \frac{k}{k_{\text{NL}}},
\end{equation}
where \( k_{\text{NL}} \) is the scale at which nonlinearities become significant, satisfying:
\begin{equation}
    4 \pi k_{\text{NL}}^3 C^\delta_{\text{lin}}(k_{\text{NL}}, 0) = 1.
\end{equation}
The coefficients \( a_i \) are:
\begin{align*}
    a_1 &= 0.484, \quad a_2 = 3.740, \quad a_3 = -0.849, \quad a_4 = 0.392, \\
    a_5 &= 1.013, \quad a_6 = -0.575, \quad a_7 = 0.128, \quad a_8 = -0.722, \quad a_9 = -0.926.
\end{align*}

\subsection{Fisher matrix analysis}
SNR and parameter constraints are obtained through a Fisher matrix analysis as is standard. A full derivation of below statements can be found in appendix \ref{sec:fisher}. The auto and cross power spectrum Fisher matrix is given as
\begin{equation*}
    F_{\alpha\beta} = \sum_{l} \sum_{XY}\sum_{X'Y'}(2l+1)\partial_\alpha C_l^{XY} (C^{-1})_l^{XX'}(C^{-1})_l^{YY'} \partial_\beta C^{X'Y'}_l,
\end{equation*}
where
\begin{equation*}
    C_{l} := \begin{pmatrix}
        C^{\psi_\kappa\psi_\kappa}_l & C^{\psi_\kappa\psi_\gamma}_l \\
        C^{\psi_\kappa\psi_\gamma}_l & C^{\psi_\gamma\psi_\gamma}_l
    \end{pmatrix}.
\end{equation*}
For the bispectra we instead get 
\begin{equation*}
    F_{\alpha\beta} = \sum_{l_1 \leq l_2 \leq l_3} \frac{\mathcal P _{l_1l_2l_3}}{6}
    \sum_{XYZ}\sum_{X'Y'Z'} 
    \partial_\alpha B^{X Y Z}_{l_1 l_2 l_3} 
    (C^{-1})^{X X'}_{l_1}
    (C^{-1})^{Y Y'}_{l_2}
    (C^{-1})^{Z Z'}_{l_3}
    \partial_\beta B^{X' Y' Z'}_{l_1 l_2 l_3},
\end{equation*}
with $\mathcal P_{l_1l_2l_3}$ defined as the number of distinct permutations that can be made with $l_1l_2l_3$. When only considering autospectra the formulas simplify to
\begin{equation*}
    F_{\alpha\beta} = \sum_{l} \sum_{XY}\sum_{X'Y'}\frac{2l+1}{2}\frac{\partial_{\alpha}C_l^{XX} \partial_\beta C^{XX}_l}{(C_l^{XX})^2},
\end{equation*}
and (for bispectra)
\begin{equation*}
    F_{\alpha\beta} = \sum_{l_1 \leq l_2 \leq l_3} \frac{\mathcal P _{l_1l_2l_3}}{6}
    \frac{\partial_\alpha B^{X X X}_{l_1 l_2 l_3} 
    \partial_\beta B^{XXX}_{l_1 l_2 l_3}}{C^{XX}_{l_1}C^{XX}_{l_2}C^{XX}_{l_3}}.
\end{equation*}

@@

% lmin, lmax, galaxy redshift distribution, fiducial values (!), noise levels, 
% for SO, CMB - S4, Euclid, LSST
We chose our fiducial cosmology in line with the results from the Planck mission \cite{planckresults}, see table \ref{tab:fiducialpars}.
\begin{table}[h!]
    \centering
    \begin{tabular}{|c|c|c|}
        \hline
        \textbf{parameter} & \textbf{notation} & \textbf{fiducial value} \\
        \hline
        Hubble constant & $H_0$ & $67.4 \, \text{km/s/Mpc}$ \\
        \hline
        Physical baryon density parameter & $\Omega_b h^2$ & $0.0223$ \\
        \hline
        Physical cold dark matter density parameter & $\Omega_c h^2$ & $0.119$\\
        \hline
        Scalar spectral index & $n_s$ & 0.965 \\
        \hline
        Amplitude of primordial scalar fluctuations & $A_s$ & $2.13 \times 10^{-9}$ \\
        \hline
        Sum of neutrino masses & $\sum m_\nu$ & $0.06$\\
        \hline
        Dark energy equation of state parameter & $w_0$ & -1 \\ \hline
    \end{tabular}
    \caption{Cosmological parameters considered ($\Lambda \text{CDM} + w_0 + m_\nu$) and their fiducial values used.}
    \label{tab:fiducialpars}
\end{table}

@@

This paper considers noise levels for ``stage 3'' and ``stage 4'' weak lensing surveys. Values can be found in table \ref{tab:noiselevels}, including the experiments they are representative of. A comparison of noise power spectrum to lensing power spectrum can be found in figure \ref{fig:lpsplusnoise}.

@@

CMB lensing noise is estimated through the quadratic estimator developed in \cite{cmblensingestimator}. We consider experiments that only measure polarization noise. This is similarly done in \cite{Namikawa_2016}. The parameters characterizing the noise levels are thus beam width, $\sigma$ and polarization white noise, $\Delta_P$. Galaxy lensing is determined by measuring lensing shear (see appendices \ref{sec:weaklensing} and \ref{sec:shear}). The noise in this measurement is dominated by scale independent shot noise and has associated noise power spectrum $N_l^{\text{shear}} = \sigma_{\text{rms}}^2 / n_g$. $n_g$ is the amount of galaxies observed per unit angle. The noise power spectrum for the lensing potential then equals
\begin{equation*}
    N_l^{\text{lens. potential}} = \frac{(l-1)l(l+1)(l+2)}{4}N_l^{\text{shear}}.
\end{equation*}
For parameter constraints we use stage 4 noise levels.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{/Users/jonasfrugte/Desktop/Research_Project/fisher_calc_weak_lensing/code/plots/spectraplusnoise.pdf}
    \caption{CMB (right) and galaxy (left) lensing potential power spectra compared to associated experimental noise. Current (stage 3) noise values are displayed as well as near future (stage 4) noise values. The CMB lensing experiment uses only polarization. CMB noise values are chosen in accordance with \cite{Namikawa_2016}. Shear noise values are chosen to be similar to e.g. Euclid measurements for stage 4 and e.g. KiDS for stage 3.}
    \label{fig:lpsplusnoise}
\end{figure}

@@

\section{Results}\label{sec:results}
Lensing power spectra are well detectable in nearly all cases. 
The signal to noise ratios of the CMB and galaxy lensing bispectra v.s. the maximum multipole moment measured can be found in figure \ref{fig:snrplots}. Shear bispectra can be measured well by both stage 3 and stage 4 experiments, while CMB lensing is only semi detectable by a stage 3 experiment but well detectable by stage 4 experiments.
%It can be seen that the CMB bispectrum signal to noise never exceeds 2 even for $l_{\max} = 2000$, meaning that this signal is not likely to be able to contribute to parameter constraints in future surveys. All other quantities should be well detectable for $l_{\max}\geq 500$.
\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/snrplots.pdf}
    \caption{Signal to noise ratios for galaxy lensing (dashed) and CMB lensing (non-dashes) power- (in red) and bispectra (in blue). The $x$-axis labels the maximum multipole moment measured up to, starting from $l_{\min}=2$.}
    \label{fig:snrplots}
\end{figure}

Parameter constraints for the full set of $\Lambda$ CDM parameters are shown in figure \ref{fig:paramconstraintsall}. The combination $H_0$, $\sigma_8$, and $\Omega_m$ is particularly well constrained. Constraints for these parameters alone are found in figure \ref{fig:paramconstraintstight}. Constraints are significantly improved by including bispectrum information in nearly all cases. This falls in line with a study in ref \cite{Namikawa_2016}, where a 33 percent and 35 percent improvement was found for $m_\nu$ and $w_0$ by measuring the CMB lensing bispectrum in addition to the powerspectrum. Constraints also improve significantly, though not quite as much, when combining CMB and galaxy lensing, compared to only measuring one of these signals. Some key observations are:
\begin{itemize}
    \item Lensing powerspectra alone are not sufficient to competitively constrain $\Lambda\mathrm{CDM} + w_0 + m_\nu$ even in the case of stage 4 surveys.
    \item For low-enough noise values (in particular for stage 4 noise) the bispectra by themselves can competitively constrain the parameters and contain more information than the power spectra. If we denote the multipole moment at which the noise surpasses the powerspectrum signal as $\ell_{N>C}$, than the fisher matrices for the bispectra roughly scale as $F^{B}_{\alpha\beta}\propto \ell_{N>C}^3$ while for the powerspectra they scale as $F^{C}_{\alpha\beta}\propto \ell_{N>C}^2$ (with an extra multipole factor due to the $2\ell + 1$ factor in the powerspectrum estimator covariance). The powerspectrum is sufficiently bigger than the bispectrum that it contains more information given current day noise levels, however as surveys get better the bispectra may eventually contain more information because of the more favorable $\ell_{N>C}$ scaling.

    \item The information from bispectra is sensitive to the cosmology in a different way than that of the powerspectra. The main parameter combinations where approximate degeneracies are broken are $(n_s,\Omega_b h^2),\;(\Omega_b h^2,H_0),\;(\Omega_c h^2,H_0),\;(\Omega_c h^2,\Omega_b h^2)\;(n_s,\Omega_c h^2),\;(m_\nu,H_0),\;(m_\nu,n_s),$ $(A_s,\Omega_b h^2)$. These means that even if stage 4 surveys surveys aren't able to reach the noise levels assumed in this paper, the bispectra will likely still lead to significantly tighter constraints.

    \item Stage 4 CMB and galaxy weak lensing surveys have similar constraining power if both measure powerspectra and bispectra. Galaxy surveys are usually somewhat better, but not significantly so. Some degeneracies are broken but not quite as much as when looking at combining power and bispectra. The main parameter combinations where this happens are  $(n_s, H_0),\;(n_s,\Omega_b h^2),\;(m_\nu,H_0),\;(m_\nu,\Omega_s),\;(A_s,H_0)\;(w_0,A_s),\;(w_0, m_\nu),\;(w_0,\Omega_s)$.
  \end{itemize}
